---
title: "Homework 1: Namit Shrivastava"
format: pdf
editor: visual
---
## 1. Select a simple random sample (SRS) of size n = 20 from this frame. Each student will select a different simple random sample, using the R code set.seed(the last four digits of your UM/UMD student ID). Note that we are simulating the notion of hypothetical repeated random sampling using the same SRS design! The class has 30 enrolled students and would generate 30 samples.

```{r}
library(readxl)
data <- read_excel("fem1524_admin.xlsx")
set.seed(5570)
n <- 20
SRS <- data[sample(nrow(data), n), ]
SRS
```

## 2. Give, in selection order, the list of the 20 four-digit selection number (IDs) and the values of AGE for the women in your sample.

```{r}
SRS_list <- SRS[, c("ID", "AGER")]
SRS_list
```

## 3. Compute the sample estimate of the mean age. What else would we need to compute (be specific) to make inference about the mean age of the population?

```{r}
mean_age <- mean(SRS_list$AGER)
mean_age
```
So to make inference, we also need the variance (specifically the standard error) of the sample mean.

## 4. What would we call the distribution that we would see if we plotted all 30 sample estimates of the mean age (computed from the 30 unique samples generated by the students in the class)? What would we call the standard deviation of this distribution?

Alright so if everyone in the class repeats this selection procedure, we will get 30 different sample means of age. The collection of those means is called the sampling distribution of the sample mean and its standard deviation will be called the standard error of the mean.


