---
title: "Homework 2: Namit Shrivastava"
format: pdf
editor: visual
---

## 1. A stratified random sample of graduates from the academic departments at a small university is to be selected to estimate the mean starting salary of graduates from that university. The university has five academic departments, and preliminary data are obtained from an attempted census in a recent year:

| Stratum | Department       | $N_h$ | $\bar{Y}_h$ |   $S_h$ |
|:--------|:-----------------|:-----:|------------:|--------:|
| 1       | Humanities       |  20   |      46,800 |  81,200 |
| 2       | Social Sciences  |  90   |      61,500 | 101,700 |
| 3       | Natural Sciences |  120  |      76,100 | 130,900 |
| 4       | Engineering      |  140  |      89,500 | 170,100 |
| 5       | Business         |  200  |      95,500 | 216,400 |

## a) Now what is the mean starting salary per graduate across all departments in the $\bar{Y}_h$ population, based on the data collected from this previous attempted census?

The overall mean salary $\bar{Y}$ is calculated as:

$$\bar{Y} = \frac{\sum_{h=1}^{H} N_h\bar{Y}_h}{\sum_{h=1}^{H} N_h}$$

$$\begin{aligned}
\bar{Y} &= \frac{(20 \times 46,800) + (90 \times 61,500) + (120 \times 76,100) + (140 \times 89,500) + (200 \times 95,500)}{20 + 90 + 120 + 140 + 200} \\
&= \frac{936,000 + 5,535,000 + 9,132,000 + 12,530,000 + 19,100,000}{570} \\
&= \frac{47,233,000}{570}
\end{aligned}
\bar{Y} = 82864.91
$$ 

The R code for this:
```{r}
# Defining the data
strata <- c(1, 2, 3, 4, 5)
departments <- c("Humanities", "Social Sciences", "Natural Sciences", 
                 "Engineering", "Business")
Nh <- c(20, 90, 120, 140, 200)
Yh_bar <- c(46800, 61500, 76100, 89500, 95500)

# Calculating Nh * Yh_bar for each stratum
Nh_Yh_bar <- Nh * Yh_bar

# Calculating the overall mean
Y_bar <- sum(Nh_Yh_bar) / sum(Nh)

cat("The mean starting salary per graduate across all departments is $", 
    round(Y_bar, 2), "\n")
```

## b) What is the average within-stratum element variance ?

The average within-stratum element variance $S^2_w$ is calculated as:

$$S^2_w = \sum_{h=1}^{H} W_h S^2_h$$

where: - $W_h = N_h/N$ is the stratum weight - $S^2_h$ is the element variance in stratum $h$ - $H = 5$ departments

$$\begin{aligned}
S^2_w &= \sum_{h=1}^{5} \frac{N_h}{N} S^2_h \\
&= \frac{20}{570}(81,200)^2 + \frac{90}{570}(101,700)^2 + \frac{120}{570}(130,900)^2 + \frac{140}{570}(170,100)^2 + \frac{200}{570}(216,400)^2 \\
&= \frac{20}{570}(6,593,440,000) + \frac{90}{570}(10,342,890,000) + \frac{120}{570}(17,134,810,000) + \\
&\quad \frac{140}{570}(28,934,010,000) + \frac{200}{570}(46,828,960,000) \\
&= 231,347,368 + 1,633,613,684 + 3,607,328,421 + 7,095,669,474 + 16,441,619,123 \\
&= 29,009,578,070
\end{aligned}
$$

Now The R code for this will be:

```{r}
# Defining the data
Nh <- c(20, 90, 120, 140, 200)
Sh <- c(81200, 101700, 130900, 170100, 216400)

# Calculating total population size
N <- sum(Nh)

# Calculating stratum weights
Wh <- Nh / N

# Calculating S^2_h
S2h <- Sh^2

# Calculating S^2_w
S2w <- sum(Wh * S2h)

cat("The average within-stratum element variance (S^2_w) is:", S2w, "\n")
```

So this value represents the weighted average of the variances within each stratum, giving us an overall measure of the variability within the strata of the population. Now doing a sqrt of the S\^2_w will give sqrt(29009578070) = \$170,321, which shows the average standard deviation across strata is about \$170,321.

## c) For a sample of n = 100, what is the proportionate allocation?

For $n = 100$, the proportionate allocation is given by:

$$n_h = n \times \frac{N_h}{N}$$

where: - $n_h$ is the sample size for stratum $h$ - $n = 100$ is the total sample size - $N_h$ is the population size of stratum $h$ - $N$ is the total population size

```{r}
# Defining the data
Nh <- c(20, 90, 120, 140, 200)
n <- 100

# Calculating total population size
N <- sum(Nh)

# Calculating proportionate allocation
n_prop <- n * (Nh / N)

print(round(n_prop, 0))
```

Hence, the proportionate allocation for n = 100 is: Humanities: 3 Social Sciences: 16 Natural Sciences: 21 Engineering: 25 Business: 35

## d) For a sample of n = 100, what is the Neyman allocation?

The Neyman allocation is given by:

$$n_h = n \times \frac{N_h S_h}{\sum_{h=1}^{H} N_h S_h}$$

where: - $n_h$ is the sample size for stratum $h$ - $n = 100$ is the total sample size - $N_h$ is the population size of stratum $h$ - $S_h$ is the standard deviation of stratum $h$ - $H = 5$ is the number of strata

```{r}
# Defining the data
Nh <- c(20, 90, 120, 140, 200)
Sh <- c(81200, 101700, 130900, 170100, 216400)
n <- 100

# Calculating Neyman allocation
Nh_Sh <- Nh * Sh
n_neyman <- n * (Nh_Sh / sum(Nh_Sh))

print(round(n_neyman, 0))
```

Hence, the Neyman allocation for n = 100 is: Humanities: 2 Social Sciences: 10 Natural Sciences: 17 Engineering: 25 Business: 46

## e) Estimate the sampling variance of the mean for the proportionate c) and Neyman d) allocations.

The sampling variance of the mean for stratified sampling is:

$$var(\bar{y}_w) = \sum_h W_h^2\frac{1-f_h}{n_h}s_h^2\text{ where }f_h=\frac{n_h}{N_h}$$

where:

W_h = N_h/N (stratum weight) n_h = allocated sample size for stratum h N_h = population size for stratum h S_h\^2 = variance in stratum h

### Step-by-step calculations:

For Proportionate Allocation:

For Humanities (h=1): $$(\frac{20}{570})^2 (1-\frac{4}{20})\frac{81,200^2}{4} = 1,623,500$$

For Social Sciences (h=2): $$(\frac{90}{570})^2 (1-\frac{16}{90})\frac{101,700^2}{16} = 13,250,932$$

For Natural Sciences (h=3): $$(\frac{120}{570})^2 (1-\frac{21}{120})\frac{130,900^2}{21} = 29,835,047$$

For Engineering (h=4): $$(\frac{140}{570})^2 (1-\frac{25}{140})\frac{170,100^2}{25} = 57,351,500$$

For Business (h=5): $$(\frac{200}{570})^2 (1-\frac{35}{200})\frac{216,400^2}{35} = 135,897,259$$

Total Proportionate Variance = 237,958,239

For Neyman Allocation:

For Humanities (h=1): $$(\frac{20}{570})^2 (1-\frac{2}{20})\frac{81,200^2}{2} = 3,652,875$$

For Social Sciences (h=2): $$(\frac{90}{570})^2 (1-\frac{10}{90})\frac{101,700^2}{10} = 22,920,532$$

For Natural Sciences (h=3): $$(\frac{120}{570})^2 (1-\frac{17}{120})\frac{130,900^2}{17} = 38,344,151$$

For Engineering (h=4): $$(\frac{140}{570})^2 (1-\frac{25}{140})\frac{170,100^2}{25} = 57,351,500$$

For Business (h=5): $$(\frac{200}{570})^2 (1-\frac{46}{200})\frac{216,400^2}{46} = 96,506,749$$

Total Neyman Variance = 218,775,808

```{r}
# Data setup
Nh <- c(20, 90, 120, 140, 200)
N <- sum(Nh)
Wh <- Nh/N
Sh2 <- c(81200^2, 101700^2, 130900^2, 170100^2, 216400^2)
n <- 100

# Proportionate allocation
n_prop <- round(n * (Nh/N))

# Calculating proportionate variance
prop_var <- sum(Wh^2 * (1 - n_prop/Nh) * Sh2/n_prop)

# Neyman allocation
n_neyman <- round(100 * (Nh * sqrt(Sh2))/sum(Nh * sqrt(Sh2)))

# Calculating Neyman variance
neyman_var <- sum(Wh^2 * (1 - n_neyman/Nh) * Sh2/n_neyman)

cat("Proportionate Variance:", prop_var, "\n")
cat("Neyman Variance:", neyman_var, "\n")
```

## f) Estimate the total element variance S\^2

The total element variance is calculated as:

$$S^2 = \sum_h \frac{N_h-1}{N-1}S_h^2+\sum_h \frac{N_h-1}{N-1}(\bar{Y}_h-\bar{Y})^2$$

Step-by-step calculation:

For Within-stratum component: $$\begin{aligned}
\sum_h \frac{N_h-1}{569}S_h^2 &= \frac{19}{569}(81,200)^2 + \frac{89}{569}(101,700)^2 + \frac{119}{569}(130,900)^2 + \\
&\quad \frac{139}{569}(170,100)^2 + \frac{199}{569}(216,400)^2
\end{aligned}
$$

For Between-stratum component: $$\begin{aligned}
\sum_h \frac{N_h-1}{569}(\bar{Y}_h-82,864.91)^2 &= \frac{19}{569}(46,800-82,864.91)^2 + \frac{89}{569}(61,500-82,864.91)^2 + \\
&\quad \frac{119}{569}(76,100-82,864.91)^2 + \frac{139}{569}(89,500-82,864.91)^2 + \\
&\quad \frac{199}{569}(95,500-82,864.91)^2
\end{aligned}
$$

```{r}
# Data setup
Nh <- c(20, 90, 120, 140, 200)
Yh_bar <- c(46800, 61500, 76100, 89500, 95500)
Sh2 <- c(81200^2, 101700^2, 130900^2, 170100^2, 216400^2)
Y_bar <- 82864.91
N <- sum(Nh)

# Calculating total element variance
S2 <- sum((Nh-1)/(N-1) * (Sh2 + (Yh_bar - Y_bar)^2))

cat("Total element variance (S^2):", S2, "\n")
cat("S^2 = ", format(S2, scientific = TRUE))
```

## g) What are the design effects of the proportionate and Neyman allocations? based on the results?

The design effect (DEFF) is calculated as:

$$deff = \frac{var(\bar{y})}{var_{SRS}(\bar{y})} = \frac{s_w^2}{s^2}$$

For Proportionate Allocation: $$deff_{prop} = \frac{237,958,239}{2,905,852,006} = 0.08188932$$

For Neyman Allocation: $$deff_{neyman} = \frac{218,775,808}{2,905,852,006} = 0.07528801$$

```{r}
# Previously calculated variances from part e
prop_var <- 237958239    # Proportionate variance
neyman_var <- 218775808  # Neyman variance

# Total element variance from part f
S2 <- 2905852006        # S^2 calculated earlier

# Calculating design effects
deff_prop <- prop_var/S2
deff_neyman <- neyman_var/S2

cat('Design Effect of Proportion Allocation:\t', deff_prop,
    '\nDesign Effect of Neyman Allocation:\t', deff_neyman)
```

Ok so the design effects for both allocations are substantially less than 1, indicating that both stratified designs are much more efficient than simple random sampling. The proportionate allocation has a design effect of about 0.082, while the Neyman allocation's design effect is about 0.075.

In essence, these low numbers indicate that the variability or error in the estimates is greatly reduced, so the study becomes more powerful and resource-efficient, with the small difference suggesting that the Neyman allocation might offer a slight advantage.

## h) Suppose that the cost-per-element was not the same in each stratum: $$C_1 = C_2 = C_3 = \$30, \text{ and } C_4 = C_5 = \$40$$

## 1) The client requesting a stratified sample design has indicated that the total available data collection budget is $$C = \$5,000$$, with the stratum- specific costs per element listed above. What allocation will minimize the sampling variance of the mean under these cost constraints?

The optimal allocation formula with cost constraints is:

$$n_h = \frac{N_h S_h/\sqrt{C_h}}{\sum_{i=1}^H N_i S_i/\sqrt{C_i}} \times \frac{C}{C_h}$$

```{r}
# Given data
Nh <- c(20, 90, 120, 140, 200)
Sh <- c(81200, 101700, 130900, 170100, 216400)
Ch <- c(30, 30, 30, 40, 40)
C <- 5000

# Calculating optimal allocation
Nh_Sh_sqrtCh <- Nh * Sh / sqrt(Ch)
n <- C / sum(Ch * Nh_Sh_sqrtCh / sum(Nh_Sh_sqrtCh))
n_opt <- n * Nh_Sh_sqrtCh / sum(Nh_Sh_sqrtCh)

# Print results
print(round(n_opt, 0))
```

Based on the optimal allocation calculation that accounts for both the stratum variability and cost constraints, I found that the sample should be distributed as: (Rounded off) 3 participants from Humanities 15 from Social Sciences 25 from Natural Sciences 33 from Engineering 60 from Business

This allocation minimizes the sampling variance of the mean starting salary while staying within the \$5,000 budget, taking into account that sampling from Engineering and Business costs more per participant (\$40) compared to the other departments (\$30).

## 2) Estimate the expected sampling variance and design effect of the mean starting salary under this allocation. (Hint: Think carefully about the definition of a design effect)

The sampling variance for stratified sampling with cost constraints is:

$$var(\bar{y}_w) = \sum_h W_h^2\frac{1-f_h}{n_h}s_h^2\text{ where }f_h=\frac{n_h}{N_h}$$

$$\begin{aligned} var(\bar{y}_w) &= \sum_h W_h^2\frac{1-f_h}{n_h}s_h^2 \ &= \sum_h (\frac{N_h}{N})^2(1-\frac{n_h}{N_h})\frac{s_h^2}{n_h} \ &= 148,362,056 \end{aligned}$$

Design Effect: $$\text{Design Effect} = \frac{var(\bar{y})}{var_{SRS}(\bar{y})} = \frac{148,362,056}{2,905,852,006} = 0.051$$

```{r}
# Data setup
Nh <- c(20, 90, 120, 140, 200)
N <- sum(Nh)
Wh <- Nh/N
Sh2 <- c(81200^2, 101700^2, 130900^2, 170100^2, 216400^2)

# Cost allocation (from previous calculation)
n_cost <- c(3, 15, 25, 33, 60)

# Calculating variance for cost allocation
cost_var <- sum(Wh^2 * (1 - n_cost/Nh) * Sh2/n_cost)

# Calculating design effect
S2 <- 2905852006  # total element variance from part f
deff_cost <- cost_var/S2

cat("Cost Allocation Variance:", cost_var, "\n")
cat("Design Effect:", deff_cost, "\n")
```

So, the results show that the cost-constrained stratified sampling design is extremely efficient, with a design effect of 0.051. This means it reduces the variance by about 95% compared to simple random sampling, which is a substantial improvement in precision.

Even with different costs per stratum and budget constraints, this allocation achieves much better precision than simple random sampling would provide.
