---
title: "Homework 3: Namit Shrivastava"
format: pdf
editor: visual
---

## 1. The following are cluster totals $$y_\alpha$$ from a = 20 clusters of exactly b = 10 women (between the ages of 15 and 24) each. These clusters and the young women were sampled from the population frame used for Homework 1. The cluster totals $$y_\alpha$$ are the number of women who have ever been pregnant. Assume that the clusters were selected at random and with replacement, and the students were selected with epsem and without replacement. The sampling fraction is f = ab/AB = n/N = 200/2,920 = 1 / 14.6, meaning that the finite population correction (fpc) should not be ignored in this case. 

| Cluster | $y_\alpha$ |
|:--------|:----------:|
| 1       | 4          |
| 2       | 4          |
| 3       | 3          |
| 4       | 6          |
| 5       | 4          |
| 6       | 6          |
| 7       | 3          |
| 8       | 4          |
| 9       | 4          |
| 10      | 1          |
| 11      | 1          |
| 12      | 8          |
| 13      | 3          |
| 14      | 3          |
| 15      | 5          |
| 16      | 6          |
| 17      | 4          |
| 18      | 5          |
| 19      | 8          |
| 20      | 5          |

## a) Compute an estimate of the mean , y bar its standard error, and a 95% confidence interval for the population mean. (Hint: the degrees of freedom used in computing this confidence interval should not be 199.)

Now for cluster sampling, the estimate of the population mean is:

$$\bar{y} = \frac{1}{ab}\sum_{\alpha=1}^a y_\alpha$$

where:
- $a$ is the number of clusters (20)
- $b$ is the cluster size (10)
- $y_\alpha$ is the cluster total

The variance of the mean is:

$$V(\bar{y}) = \frac{1}{a^2b^2}\sum_{\alpha=1}^a(y_\alpha - \bar{y}b)^2/(a-1)$$

And the standard error is:

$$SE(\bar{y}) = \sqrt{V(\bar{y})}$$

So step by step calculation will be:

Mean: $$\bar{y} = \frac{87}{(20 \times 10)} = 0.435$$

Variance: $$V(\bar{y}) = \frac{1}{20^2}\sum_{\alpha=1}^{20}(y_\alpha/10 - 0.435)^2/19$$

Standard Error: $$SE(\bar{y}) = \sqrt{V(\bar{y})}$$

95% Confidence Interval: $$\bar{y} \pm t_{19,0.975} \times SE(\bar{y})$$

```{r}
# Data
y_alpha <- c(4, 4, 3, 6, 4, 6, 3, 4, 4, 1, 1, 8, 3, 3, 5, 6, 4, 5, 8, 5)
a <- 20  # number of clusters
b <- 10  # cluster size
N <- 2920  # total population size
n <- a * b  # sample size

# Calculating mean
y_bar <- sum(y_alpha)/(a*b)

# Calculating variance and SE with fpc
cluster_means <- y_alpha/b
var_y <- var(cluster_means)  # using a-1 degrees of freedom
fpc <- sqrt((N - n) / N)
se_y <- sqrt(var_y/a) * fpc

# Calculating 95% CI using t-distribution with a-1 df
t_value <- qt(0.975, df = a-1)
ci_lower <- y_bar - t_value * se_y
ci_upper <- y_bar + t_value * se_y

cat("Mean:", y_bar, "\n")
cat("Standard Error:", se_y, "\n")
cat("95% CI:", c(ci_lower, ci_upper), "\n")
```
Based on the analysis of the cluster sampling data, I found that the estimated mean number of women who have ever been pregnant in each cluster of 10 women is approximately 4.35. 

The standard error of this estimate is about 0.040, which gives an idea of the variability of the sample mean. 

Using this standard error, I calculated a 95% confidence interval for the population mean, which ranges from approximately 3.50 to 5.20 women per cluster. This means one can be 95% confident that the true mean number of women who have ever been pregnant in each cluster of 10 women falls within this interval.

Also the relatively narrow confidence interval suggests that the estimate is fairly precise, even when accounting for the finite population correction. This precision indicates that the sampling method was effective in capturing the variability within the population.


## b) Estimate the standard error of the mean that you would expect if the sample consisted of a = 40 clusters of b = 10 each. (Hint: What about this design has not changed, and what quantity needed to answer this question could therefore be considered portable?)

For cluster sampling with a different number of clusters, I can use the fact that the cluster means variance remains constant. The new standard error would be:

$$SE_{new}(\bar{y}) = \sqrt{\frac{V(y_\alpha/b)}{a_{new}}} \times \sqrt{\frac{N-n_{new}}{N}}$$

where:
- $V(y_\alpha/b)$ is the variance of cluster means (portable from previous calculation)
- $a_{new} = 40$ is the new number of clusters
- $n_{new} = a_{new} \times b$ is the new total sample size
- $N = 2920$ is the total population size

So a step by step calculation will be:

Using portable variance of cluster means: $$V(y_\alpha/b) = 0.0326$$

Calculating new standard error: $$SE_{new} = \sqrt{\frac{0.0326}{40}} \times \sqrt{\frac{2920-400}{2920}} = 0.0267$$

```{r}
# Taking Data from previous calculation
var_cluster_means <- var(y_alpha/b)  # portable variance of cluster means
a_new <- 40  # new number of clusters
b <- 10      # cluster size (unchanged)
N <- 2920    # total population size
n_new <- a_new * b  # new sample size

# Calculating new standard error
fpc_new <- sqrt((N - n_new)/N)
se_new <- sqrt(var_cluster_means/a_new) * fpc_new

cat("Expected Standard Error with 40 clusters:", se_new, "\n")
```

Ok so looking at the results I got, I found that if one is to double the number of clusters from 20 to 40 while keeping the cluster size at 10 women each, the expected standard error would decrease to approximately 0.027. This is notably smaller than the original standard error of 0.040, which makes sense because one is sampling more clusters. 

The reduction in standard error suggests that increasing the number of clusters would improve the precision of the estimate by about 32%. This improvement in precision demonstrates how sampling more clusters, even while maintaining the same cluster size, can lead to better estimates of the population mean number of pregnant women per cluster.
